---
title: "Analyse2_with_ADMI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyse2_with_ADMI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Analysis sketch - this version uses analyse and pool functions in RBMI
based on the example in the vignette https://insightsengineering.github.io/rbmi/main/articles/advanced.html

The analysis is as follows:
Rubin’s Rules after standardization: Following multiple imputation, we would fit 1000 logistic regression models and for each, obtain the marginal estimates using beeca. We would then pool these 1000 beeca outputs using Rubin’s Rules.


```{r setup}
# Load libraries
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(rbmi)
library(beeca)
library(rbmiUtils)

# check if rstan is installed localy and install if not 
# if (!rlang::is_installed("rstan")) {
#   install.packages("rstan", repos = "https://cloud.r-project.org/")
#   message("rstan has been successfully installed.")
# } else {
#   message("rstan is already installed.")
# }


# set seed 
set.seed(1974)


## Define global rbmi constants - define imputation parameters
IMPUTATIONS <- 100
WARMUP <- 200
THIN <- 5


# load example data
data("ADEFF")


# Prepare the data
ADEFF$TRT <- factor(ADEFF$TRT01P, levels = c("Placebo", "Drug A"))
ADEFF$USUBJID <- factor(ADEFF$USUBJID)
ADEFF$AVISIT <- factor(ADEFF$AVISIT)
```

Define function to compare last visit using a responder analysis

```{r gcomp}
## function to compare last visit using a responder analysis
gcomp_responder <- function(data, ...) {

    ## fit the GLM model
    model <- glm(CRIT1FLN ~ TRT + BASE + STRATA + REGION, data = data, family = binomial)

    ## Calculate the marginal treatment effect estimate and associated variance
    ## for a difference contrast using the Ge et al. method
    ## with robust HC0 sandwich variance
    marginal_fit <- get_marginal_effect(model,
                                        trt = "TRT",
                                        method = "Ge",
                                        type = "HC0",
                                        contrast = "diff",
                                        reference = "Placebo")

    ## Extract results
    marginal_results <- marginal_fit$marginal_results
    diff_est <- marginal_results[marginal_results$STAT == "diff", "STATVAL"][[1]]
    diff_se <- marginal_results[marginal_results$STAT == "diff_se", "STATVAL"][[1]]

    # Return a named list with the estimate and standard error
    res <- list(
        trt = list(
            est = diff_est,
            se = diff_se,
            df = NA
        )
    )
    return(res)
}

```

# STEP 1: Impute missing data


```{r define}
# Define key variables for rbmi
vars <- rbmi::set_vars(
    subjid = "USUBJID",
    visit = "AVISIT",
    group = "TRT",
    outcome = "CHG",
    covariates = c("BASE", "STRATA", "REGION")
)

# Define which imputation method to use 
# (here: Bayesian multiple imputation with 200 imputed datsets)
method <- rbmi::method_bayes(
    n_samples = 100,
    control = rbmi::control_bayes(
      warmup = 200,
      thin = 2
    )
  )

# Create a dataset with the necessary columns
dat <- ADEFF |>
    dplyr::select(USUBJID, STRATA, REGION, REGIONC, TRT, BASE, CHG, AVISIT)

# Fit the imputation model
draws_obj <- rbmi::draws(
    data = dat,
    vars = vars,
    method = method,
    quiet = FALSE
)

# Impute the datasets
impute_obj <- rbmi::impute(
    draws_obj,
    references = c("Placebo" = "Placebo", "Drug A" = "Placebo")
)


## get the imputed datq as a data frame
ADMI <- get_imputed_data(impute_obj)
```


Transform the imputed data to include additional derived variables. 

For example, add responder criteria based on the outcome: We convert the 
continuous version of the endpoint into binary version of the endpoint 
(e.g. if the endpoint>3, then it is a responder; if it is <=3, then it 
is a non-responder).


```{r get_data}
ADMI <- ADMI |>
    dplyr::mutate(
        CRIT1FLN = ifelse(CHG > 3, 1, 0), 
        CRIT1FL = ifelse(CRIT1FLN == 1, "Y", "N"),
        CRIT = "CHG > 3"
    )

# save the imputed data
#write_csv(ADMI, "ADMI.csv")

```



# STEP 2: Analyze each imputed dataset

## ANCOVA analysis

```{r}
ana_obj_ancova <- analyse_mi_data(data = ADMI,
                         vars = vars,
                         method = method,
                         fun = ancova,
                         delta = NULL)
```


### STEP 3: Pool analyzes and apply Rubin's Rules

```{r}
pool_obj_ancova <- pool(ana_obj_ancova)
```

### STEP 4: Report analysis

```{r}
pool_obj_ancova
```


## Responder analysis 

Note the vars may change from imputation to analysis

Here we focus on the new derived variable CRIT1, CRIT1FL, CRIT1FLN 

```{r}

## note the vars may change from imputation to analysis
vars <- set_vars(
    subjid = "USUBJID",
    visit = "AVISIT",
    group = "TRT",
    outcome = "CRIT1FLN",
    covariates = c("BASE", "STRATA", "REGION")
)
```

### Analyse the imputed datasets

Need to do the data transform in the fun

```{r}
ana_obj_prop2 <- analyse_mi_data(data = ADMI,
                         vars = vars,
                         method = method,
                         fun = gcomp_responder,
                         delta = NULL)
```


### STEP 3: Pool analyzes and apply Rubin's Rules

```{r}
pool_obj_prop <- pool(ana_obj_prop2)
```

### STEP 4: Report analysis

```{r}
pool_obj_prop
```
