---
title: "WIP: Delta Analysis of Imputed Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{delta_analyse_mi_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates the process of analyzing imputed data using the `rbmi` package, specifically focusing on applying a delta adjustment to missing values and running a tipping point analysis.

### Prerequisites

```{r setup}
library(dplyr)
library(rbmi)
library(rbmiUtils)
library(ggplot2)
library(purrr)

## Define global rbmi constants - define imputation parameters
IMPUTATIONS <- 200
WARMUP <- 200
THIN <- 5
SEED <- 1974


# load example data
data("ADEFF")

# set seed 
set.seed(mi_parameters$SEED)


# Prepare the data
ADEFF <- ADEFF %>%
  mutate(
    TRT = factor(TRT01P, levels = c("Placebo", "Drug A")),
    USUBJID = factor(USUBJID),
    AVISIT = factor(AVISIT)
  )
```

# Step 1: Impute Missing Data

In this step, we set up the necessary variables and impute missing data using Bayesian multiple imputation.

```{r}
# Define key variables for rbmi
vars <- rbmi::set_vars(
    subjid = "USUBJID",
    visit = "AVISIT",
    group = "TRT",
    outcome = "CHG",
    covariates = c("BASE", "STRATA", "REGION")
)

# Define which imputation method to use 
# (here: Bayesian multiple imputation with 200 imputed datsets)
method <- rbmi::method_bayes(
    n_samples = IMPUTATIONS,
    control = rbmi::control_bayes(
    warmup = WARMUP,
    thin = THIN
    )
  )


# Subset relevant columns
dat <- ADEFF %>%
    select(USUBJID, STRATA, REGION, REGIONC, TRT, BASE, CHG, AVISIT)

# Fit the imputation model and perform imputation
draws_obj <- rbmi::draws(data = dat, vars = vars, method = method, quiet = TRUE)
impute_obj <- rbmi::impute(draws_obj, references = c("Placebo" = "Placebo", "Drug A" = "Placebo"))

# Get the imputed data as a data frame
ADMI <- get_imputed_data(impute_obj)
```

# Step 2: Delta Adjustment

We now apply a delta adjustment to imputed outcomes, setting a delta value of 0.5 for all missing outcomes.

```{r}
# Create delta adjustment template
dat_delta <- delta_template(imputations = impute_obj) %>%
  mutate(delta = is_missing * 0.5)

glimpse(dat_delta)
```

# Step 3: Analyze Each Imputed Dataset

Here, we perform an ANCOVA analysis on the imputed datasets after applying the delta adjustment.

```{r}
ana_obj_ancova <- analyse_mi_data(
  data = ADMI,
  vars = vars,
  method = method,
  fun = ancova,
  delta = dat_delta
)
```

# Step 4: Pool the Results

We pool the results of the ANCOVA analyses across all imputed datasets using Rubin's Rules.

```{r}
pool_obj_ancova <- pool(ana_obj_ancova)
pool_obj_ancova
```

# Tipping Point Analysis

Next, we conduct a tipping point analysis to assess the sensitivity of the results to various delta offsets for the control and intervention groups.

WIP: the values do not make scientific sense and are to illustrate a difference. 

```{r}
perform_tipp_analysis <- function(delta_control, delta_intervention) {
  delta_df <- delta_df_init %>%
    mutate(
      delta_ctl = (TRT == "Placebo") * is_missing * delta_control,
      delta_int = (TRT == "Drug A") * is_missing * delta_intervention,
      delta = delta_ctl + delta_int
    )

  ana_obj_delta <- analyse_mi_data(
    data = ADMI %>% filter(AVISIT == "Week 48"),
    vars = vars,
    method = method,
    fun = ancova,
    delta = delta_df
  )
  
  pool_delta <- as.data.frame(pool(ana_obj_delta)) %>%
    filter(parameter == "trt_Week 48")

  list(
    trt_effect_6 = pool_delta[["est"]],
    pval_6 = pool_delta[["pval"]]
  )
}

# Initial delta template
delta_df_init <- delta_template(impute_obj) %>%
  filter(AVISIT == "Week 48")

# # Define delta ranges for control and intervention groups
# tipp_frame_grid <- expand.grid(
#   delta_control = seq(-10, 10, by = 1),
#   delta_intervention = seq(-10, 10, by = 1)
# ) %>%
#   as_tibble()
# 
# # Apply tipping point analysis across the grid
# tipp_frame <- tipp_frame_grid %>%
#   mutate(
#     results_list = map2(delta_control, delta_intervention, perform_tipp_analysis),
#     trt_effect_6 = map_dbl(results_list, "trt_effect_6"),
#     pval_6 = map_dbl(results_list, "pval_6")
#   ) %>%
#   select(-results_list) %>%
#   mutate(
#     pval = cut(
#       pval_6,
#       c(0, 0.001, 0.01, 0.05, 0.2, 1),
#       right = FALSE,
#       labels = c("<0.001", "0.001 - <0.01", "0.01- <0.05", "0.05 - <0.20", ">= 0.20")
#     )
#   )
# 
# # Display delta values leading to non-significant results
# tipp_frame %>%
#   filter(pval_6 >= 0.05)
# 
# # Plot the tipping point results
# ggplot(tipp_frame, aes(delta_control, delta_intervention, fill = pval)) +
#   geom_raster() +
#   scale_fill_manual(values = c("darkgreen", "lightgreen", "lightyellow", "orange", "red"))
```
